How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in \textit{Emys marmorata} (Testudinoidea, Emydidae).

### 2013-08-12 documents/turtle_paper.Rnw
background section updated to reflect ken comments. draft finished.

% past work on automatic taxon identification and older approaches to classifying taxa
% why use machine learning methods
\subsection{Background and system}
Differences in morphological variation between different classes has previously been analyzed using methods like linear discriminate analysis and canonical variates analysis \citep{Zelditch2004,Mitteroecker2011,Polly2007a,Polly2003,Gunduz2007,Gaubert2005b,Demandt2009}. These methods are comparatively straight forward ways of understanding the differences in morphology between classes. Also, they are very visual methods which aides with the interpretation and presentation of information. These previous studies, however, do not compare which amongst a set of candidate classification hypotheses is better. For example, studies such as those of \citet{Caumul2005a} and \citet{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme, instead of comparing the fidelity of one aspect of morphology to multiple classification schemes. Of note, however, is the work of \citet{Cardini2009a}, which compared morphological variation in marmots at both population, regional, and species levels to determine fidelity between shape each of these different hierarchical levels. Importantly, however, is that the classification models have not been generalized to testing data and training data accuracy is used almost exclusively as the metric off classification strength.

Here, we used multiple machine learning methods, both unsupervised and supervised, in order to compare different classification hypotheses. These methods provide different and unique advantages for understanding how to classify taxa, with what accuracy, and what these classifications are based on. While machine learning methods such as neural networks have been applied to studying shape variation \citep{MacLeod2007}, they have been primarily applied in the context of automated taxon identification and not in terms of group classification and strength of classification. Additionally, we investigate variation in continuous traits and not discrete differences between each class, instead focusing on differences in the multivariate quantification of shape.

% this paragraph kind of sucks
The two major classes of machine learning methods, unsupervised and supervised, are essentially extensions of known statistical methods \citep{Hastie2009}. Unsupervised learning methods are analogous to clustering and density estimation methods \citep{Kaufman1990}, while supervised learning methods are analogous classification and regression models \citep{Breiman1984}. In both cases, many of these methods are not fit via maximum likelihood and are supplemented by randomization, sorting, and partitioning algorithms along with the maximization or minimization of summary statistics in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}. The application of the alternative approaches used in this study illustrates only a sampling of the various previously derived methods for clustering observations and fitting classification models. 
% Applications of these methods outside of this study
% epidemeology, computer science, marketing
Additionally, instead of pure classification accuracy, here we use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified (see Methods).

>>>>

% past work on automatic taxon identification and older approaches to classifying taxa
% why use machine learning methods
\subsection{Background and system}
Machine learning is in many respects just an extension of known statistical methodology \citep{Hastie2009} where the emphasis is placed on inferring rules and properties of data in order to explain the underlying structure. The basic statistical mechanics are supplemented by randomization, sorting, and partitioning algorithms and along with the maximization or minimization of summary statistics in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}. Machine learning approaches have found use in medical research, epidemiology, economics and automated image identification such as handwritten zip codes \citep{Hastie2009}. Two major classes of machine learning methods are unsupervised and supervised learning. Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated and are analogous to clustering and density estimation methods \citep{Kaufman1990}. Supervised learning methods are used with labeled data where the final output of data is known and the rules for going from input to output are inferred. These are analogous classification and regression models \citep{Breiman1984}. The application of the alternative approaches used in this study illustrates only a sampling of the various previously derived methods for clustering observations and fitting classification models.

Differences in morphological variation between different classes has previously been analyzed using methods like linear discriminate analysis and canonical variates analysis \citep{Zelditch2004,Mitteroecker2011,Polly2007a,Polly2003,Gunduz2007,Gaubert2005b,Demandt2009}. These methods are comparatively straight forward ways of understanding the differences in morphology between classes. Also, they are very visual methods which aides with the interpretation and presentation of information. Previous studies, however, normally do not compare which amongst a set of candidate classification hypotheses is better. For example, studies such as those of \citet{Caumul2005a} and \citet{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme, instead of comparing the fidelity of one aspect of morphology to multiple classification schemes. Of note, however, is the work of \citet{Cardini2009a}, which compared morphological variation in marmots at both population, regional, and species levels to determine fidelity between shape each of these different hierarchical levels. Importantly, however, is that the classification models have not been generalized to testing data and training data accuracy is used almost exclusively as the metric off classification strength.

Here, we used multiple machine learning methods, both unsupervised and supervised, in order to compare different classification hypotheses. These methods provide different and unique advantages for understanding how to classify taxa, with what accuracy, and what these classifications are based on. While machine learning methods such as neural networks have been applied to studying shape variation \citep{MacLeod2007}, they have been primarily applied in the context of automated taxon identification and not in terms of group classification and strength of classification. Additionally, we investigate variation in continuous traits and not discrete differences between each class, instead focusing on differences in the multivariate quantification of shape.
Also, instead of pure classification accuracy, here we used a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified (see Methods).
